---
layout: post
section-type: post
title: Plot No 1 - smoothScatter
category: graphics
tags: [R, ggplot2]
---

Why do I write this post?
-------------------------
The motivation for this plot is the function: `graphics::smoothScatter`. So I wondered if I could imitate the main features of the plot with [ggplot2](http://ggplot2.org/).

Reproducing the plot:
---------------------
First some data:

```{r}
library(ggplot2)
library(MASS)
library(grid)
set.seed(2)
dat <- data.frame(mvrnorm(n=1000, mu=c(0, 0), Sigma=matrix(rep(c(1, 0.2), 2), nrow = 2, ncol = 2)))
names(dat) <- c("x", "y")
```

So the idea of `smoothScatter` is to present a scatter plot, basically, but wherever too many points overlap, instead of points a two dimensional density estimation is presented. This is nice especially in the case of 'a lot' of observations.

```{r fig.height=5, fig.width=10, dpi=100}
par(mfrow=c(1,2))
plot(dat$x, dat$y)
smoothScatter(dat$x, dat$y)
```
OK, very pretty, lets reproduce this feature in `ggplot2`. First thing is to add the necessary layers, which I already mentioned is a two dimensional density estimation, combined with the `geom` called 'tile'. Also I use the fill aesthetic to add color and a different palette:

```{r fig.height=4, fig.width=8, dpi=100, tidy=FALSE}
ggplot(data = dat, aes(x, y)) +
  stat_density2d(aes(fill = ..density..), geom = "tile", contour = F, n = 200) +
  scale_fill_continuous(low = "white", high = "dodgerblue4")
```
I'd like to add two more layers. One is a simple scatter plot. `ggplot2` provides a nice aesthetic, `alpha`, which controls the transparency of the `geom`. In this case I set it on `1/10` which is sufficient for `n = 1000`. The other layer is a geom called `rug`, which is another nice addition to scatter-plot-like figures:

```{r fig.height=4, fig.width=8, dpi=100, tidy=FALSE}
last_plot() +
  geom_rug(alpha = 1/10, sides = "b") +
  geom_point(alpha = 1/10)
```

Preparing it for this blog
--------------------------
The last step is to tweak the `theme`-elements, because I am interested in a nice 'picture' for this blog, I get rid of everything informative. Starting off from a standard theme, `theme_classic`, which comes close to where I want to get, I get rid of all labels, axis and the legend. Plus the `plot.margin` will be set to zero, which is nice for the presentation. A last thing is to zoom in by modifying the limits of `coord_cartesian`:

```{r fig.height=4, fig.width=8, dpi=100, tidy=FALSE}
last_plot() +
  theme_classic() +
  theme(legend.position = "none", axis.line = element_blank(), axis.ticks = element_blank(), axis.text = element_blank(), text = element_blank(), plot.margin = unit(c(0, 0, -0.56, -1), "cm")) +
  coord_cartesian(xlim=c(-2,2), ylim = c(-1, 1))
```

The very last thing is to save the plot in the correct format for display:

```{r, eval=FALSE}
ggsave("../images/plotNo1.jpg", plot = last_plot(), width = 54.19, height = 13.55, dpi = 300, units = "cm")
```
And thats it, a nice picture which used to be a statistical graph.

Other implementations
---------------------
I just found [this question on StackOverflow](http://stackoverflow.com/questions/13094827/how-to-reproduce-smoothscatters-outlier-plotting-in-ggplot) which I prefere over the above.
