---
layout: post
title: "'Functional Programming with Scala' with R"
author: "Sebastian"
tags: [R, fp, oop, aoos]
bibliography:
output: wahaniMiscs:::pdf_memo
math: true
archive: true
---

# Higher Order Functions

A function to represent:

\\[ \sum_{i = a}^b f(i) \\]

```{r}
sum <- function(f, a, b) {
  loop <- function(a, acc) {
    if (a > b) acc
    else loop(a + 1, f(a) + acc)
  }
  loop(a, 0)
}

sum(function(x) x, 1, 5) # identity
sum(function(x) x^3, 1, 5) # cube
sum(function(x) if(x == 0) 1 else x * Recall(x - 1), 1, 5) # factorial
```

# Currying

Somehow get rid of the repetition in the 3 above examples! In each line the arguments a and b are passed, although they are the same. Not necessary!

```{r}
sum <- function(f) {
  function(a, b) {
    if (a > b) 0
    else f(a) + Recall(a + 1, b)
  }
}

sum(function(x) x)(1, 5)
sum(function(x) x^3)(1, 5)
sum(function(x) if(x == 0) 1 else x * Recall(x - 1))(1, 5)
```

I don't see where we get more expressive here. The example is not convincing for me. The *real* advantage to redefine `sum` is that you know can pass `sum(function(x) x)` as an object to another function and `a` and `b` are derived in another context (scope/environment).

## Excercise

- write a product function
- write factorial in terms of product
- abstract sum/product to some other higher order function

```{r}
product <- function(f) {
  function(a, b) {
    if (a > b) 1 # change to 1
    else f(a) * Recall(a + 1, b) # change to *
  }
}

identity <- function(x) x
cube <- function(x) x^3

product(identity)(1, 10) == factorial(10)
fact <- functional::Curry(product(identity), a = 1)
fact(5)
product(cube)(1, 10) # == large number...
```

Abstract this `sum`/`product` to something more general:

```{r}
mapReduce <- function(f, combine, zero) {
  function(a, b) {
    if (a > b) zero
    else combine(f(a), Recall(a + 1, b))
  }
}

mapReduce(identity, `*`, 1)(1, 5)
mapReduce(identity, `+`, 0)(1, 5)
```

Why is this `mapReduce` different from the `Reduce` function in R? Hm, in `mapReduce` we have a recursive component where I do not see a direct corespondence in the `Reduce` function which, in contrast, reduces a given vector of values.

# Functions and Data

```{r}
devtools::install_github("wahani/aoos")
library(aoos)
Rational <- Class({
  Class <- "Rational"
  contains <- "Private"

  numer <- "numeric"
  denom <- "numeric"
  .g <- "numeric"

  .gcd <- function(a, b) if(b == 0) a else Recall(b, a %% b)

  initialize <- function(numer, denom) {
    .self$.g <- .gcd(numer, denom)
    .self$numer <- numer / .g
    .self$denom <- denom / .g

  }

  show <- function() {
    cat(paste0(.self$numer, "/", .self$denom, "\n"))
  }

  add <- function(that) {
    Rational(numer = numer * that$denom + that$numer * denom,
             denom = denom * that$denom)
  }

  neg <- function() {
    Rational(numer = -.self$numer,
             denom = .self$denom)
  }

  sub <- function(that) {
    add(that$neg())
  }

})

rational <- Rational(2, 3)
rational$add(rational)
rational$neg()
rational$sub(rational)

x <- Rational(numer = 1, denom = 3)
y <- Rational(numer = 5, denom = 7)
z <- Rational(numer = 3, denom = 2)

x$sub(y)$sub(z)

```
