---
layout: post
title: PlotNo2 - Wes Anderson style
description: "Some graphics in R using Wes Anderson colour palettes."
comments: true
categories: [R, graphics]
archive: true
---

Here I am playing around with different colour palettes and `geom_tile()` from [ggplot2](http://ggplot2.org/). I have nothing special in mind by doing the following graphs - so don't expect anything.

## Resources on colour palettes

  - [Choosing Colour Palettes Part I](http://blog.ggplot2.org/post/23995319650/choosing-colour-palettes-part-i-introduction)
  - [Choosing Colour Palettes Part II](httphttp://blog.ggplot2.org/post/24607351280/choosing-colour-palettes-part-ii-educated-choices)

## Resources on Wes Anderson palettes

  *  on the [Rovolution Analytics blog](http://blog.revolutionanalytics.com/2014/03/give-your-r-charts-that-wes-anderson-style.html)  
  *  on a blog called [Satisfaction](http://statisfaction.wordpress.com/2014/03/31/moustache-target-distribution-and-wes-anderson/)  
  *  as far as I understand, all palettes are inspired by the [WesAndersonPalettes blog](http://wesandersonpalettes.tumblr.com/)  
  *  the [Github repository of the package](https://github.com/karthik/wesanderson) - also has some examples


## Loading Packages

```{r}
library(MASS)
library(ggplot2)
library(grid)

# require(devtools)
# install_github("wesanderson", "karthik")
# Maybe this one is on CRAN by now ...
library(wesanderson)

```

## Plotting Idea 1

### Some Data

For the first plot I use a bivariate uniform distribution - the variates are independent. So this will just result in randomly-coloured tiles:

```{r}
n <- 10000
dat <- data.frame(V1 = runif(n, 0, 10), V2 = runif(n, 0, 10))
dat$fac1 <- cut(dat$V1, breaks = seq(from=0, to=10, length.out=40))
dat$fac2 <- cut(dat$V2, breaks = seq(from=0, to=10, length.out=40))
```

### Some Plot

```{r}
# Data and aesthetics:
ggplot(dat, aes(x = fac1, y = fac2,
                fill = cut(..count.., breaks = c(-1, 1, 2, 5, 10000)))) +
  # geoms:
  geom_tile(stat = "bin2d") +
  # colours:
  #scale_fill_brewer(type = "qual", palette = "Set3") +
  scale_fill_manual(values = wes_palette("Chevalier", 4)) +
  # theme:
  theme_classic() +
  theme(legend.position = "none",
        axis.line = element_blank(), axis.ticks = element_blank(),
        axis.text = element_blank(), text = element_blank())

```

## Plotting Idea 2

### Some Data

For the first plot I use a uniformly distributed and a exponentially distributed variate. Just for the visual effect.

```{r}
dat <- data.frame(V1 = runif(n, 0, 10), V2 = rexp(n, 0.5))
dat$fac1 <- cut(dat$V1, breaks = seq(from=0, to=10, length.out=40))
dat$fac2 <- cut(dat$V2, breaks = seq(from=min(dat$V2), to=max(dat$V2), length.out=100))
```

### Some Plot
```{r}
# Data and aesthetics:
ggplot(dat, aes(x = fac2, y = fac1,
                fill = ..count..)) +
  # geom:s
  geom_tile(stat = "bin2d") +
  # colours:
  #scale_fill_gradient(low = "#56B1F7", high = "#132B43") +
  scale_fill_gradientn(colours = wes_palette("Chevalier", 4)[c(2, 4, 1)]) +
  # theme:
  theme_classic() +
  coord_cartesian(xlim = c(0, 50)) +
  theme(legend.position = "none", axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        text = element_blank(),
        panel.background = element_rect(fill = wes_palette("Chevalier", 4)[3]),
        plot.margin = unit(c(0, 0, -0.57, -1), "cm"))
```
