---
layout: post
title: "The Matrix package"
author: "Sebastian"
categories: [R]
---

# Some considerations on perfomance:

This is the example from

```{r eval=FALSE}
vignette("Comparisons", package = "Matrix")
```

## Base R
```{r}
library("Matrix")
data(KNex, package = "Matrix")
y <- KNex$y
mm <- as(KNex$mm, "matrix")
system.time(naive.sol <- solve(t(mm) %*% mm) %*% t(mm) %*% y)
system.time(cpod.sol <- solve(crossprod(mm), crossprod(mm,y)))
system.time(t(mm) %*% mm)
```

## Using Matrix

This is equivalent to use `cpod.sol`:
```{r}
mm <- as(KNex$mm, "dgeMatrix")
system.time(Mat.sol <- solve(crossprod(mm), crossprod(mm, y)))
```

There is also a concept of memoization implemented as allustrated as:

```{r}
xpx <- crossprod(mm)
xpy <- crossprod(mm, y)
system.time(solve(xpx, xpy))
system.time(solve(xpx, xpy))
```

I don't know where the results are stored exactly. The document says with the original object, so either as attribute to `xpx` or `xpy`.

```{r}
names(xpx@factors)
names(xpy@factors)
```

Okay, seems that only the first argument to `solve` is modified. But in nested calls this effect is not used:

```{r}
system.time(solve(crossprod(mm), crossprod(mm, y)))
class(mm)
class(xpx)
```

Maybe the reason for this is that `crossprod` will change the data, so storing results of the cholesky decomposition would not be meaningfull as it only makes sense for the cross product and not `mm`. Still this means that creating `xpy` is not necessary:

```{r}
xpx <- crossprod(mm)
system.time(solve(xpx, crossprod(mm, y)))
system.time(solve(xpx, crossprod(mm, y)))
```

It's kind of the *manual* memoization of results you don't want to recompute.

## Taking advantage of sparse

```{r}
mm <- KNex$mm
class(mm)
```

The matrix package privides a class for sparse matrices. The gain in speed is obvious, no tricky memoization needed as a lot of elements in `mm` are indeed zero.

```{r}
system.time(sparse.sol <- solve(crossprod(mm), crossprod(mm, y)))
```

The methods seem to be polimorphic, so the return value can have different classes. `Matrix` for example will try to determine the class on it's own:

```{r}
dat <- rnorm(1e6)
dat[1:300000] <- 0
m <- Matrix(dat, ncol = 1e3, nrow = 1e2)
class(m)
```

```{r}
oldM <- matrix(rnorm(1e6), ncol = 1e3, nrow = 1e2)
oldMm <- crossprod(oldM)
class(Matrix(oldMm))
```

So this is nice!
